---
title: "Time Series Analysis - Project"
author: "Divya Chandrashekar"
date: "April 25, 2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error =TRUE)

```


```{r}
install.packages("astsa")
install.packages("devtools")
devtools::install_github("nickpoison/astsa")

library(astsa)

install.packages("Rcpp")

install.packages("acf2")
library(acf2)

install.packages("sarima")
library(sarima)

install.packages("tseries")
library(tseries)

install.packages("Rcpp")
library(Rcpp)
```

##1. Data Description

a) Data Source:U.S. Unemployment. Astsa package. Dataset - unemp
 
```{r unemp}
unemp<-astsa::unemp
```

b) Variable Description

Monthly U.S. Unemployment series (1948-1978, n = 372)
```{r}
summary(unemp)
#checking class to confirm it is a ts
class(unemp)
```

c) ts description

```{r}
start(unemp)

end(unemp)

frequency(unemp)
```
From these results it can be see that the series starts with 1948 January and ends at 1978 December, with a monthy frequency


##2. Data Exploration

a) Plotting data

```{r message=FALSE, echo=FALSE}
plot(unemp, ylab = "Unemployment",main="US Unemployment over time")
```


```{r message=FALSE, echo=FALSE}
plot(unemp,ylab = "Unemployment",main="US Unemployment over time (with trend line")
abline(reg=lm(unemp~time(unemp))) # fit a trend line
```

For further EDA,
Box plot across months to explore seasonal effects 
(Also aggregating annualy)

```{r message=FALSE, echo=FALSE}
plot(aggregate(unemp), ylab = "Unemplyment Annual",main="Annal US Unemployment over time")
boxplot(unemp ~ cycle(unemp),ylab="US Unemployment",main="Boxplot of Unemployment across months")
```


From the preliminary plot of raw data, an increasing trend can be observed.
From the boxplot across months, highest unemployment is observed in June. 


b)Transforming data

Differencing the unemployment data and plotting it
```{r message=FALSE, echo=FALSE}
d_unemp <- diff(unemp)
plot(d_unemp)
```

Seasonally differencing d_unemp and plotting it
```{r}
dd_unemp <- diff(d_unemp, lag = 12)  
plot(dd_unemp)
```


By differencing, the trend and seasonal variation in unemployment have been removed.
The series now appears to be stationary.

##3. Data Decomposition

Decomposing series into trend, seasonal and residual components
and plotting these

```{r message=FALSE, echo=FALSE}
plot(decompose(unemp))
u<-decompose(unemp)

trend<-u$trend
seasonal<-u$seasonal
residual<-u$random


plot(trend,ylab="US Unemployment",main="Plot of Unemployment trend")
plot(seasonal,ylab="US Unemployment",main="Plot of Unemployment seasonal")
plot(residual,ylab="US Unemployment",main="Plot of Unemployment residuals")

ts.plot(cbind(trend, trend+seasonal), lty=1:2)
```

Since the series shows an increasing trend, but seasonal effect does not, an additive decomposition model is appropriate.
The residuals seem to show an increasing trend, so a log transformation can be considered.

##4.Regression

Regressing unemployment data on time
```{r}
fit <- lm(unemp~time(unemp), na.action=NULL)

summary(fit)

summary(aov(fit))
num = length(unemp)


AIC(fit)/num - log(2*pi) # AIC
BIC(fit)/num - log(2*pi) # BIC
(AICc = log(sum(resid(fit)^2)/num) + (num+5)/(num-5-2)) # AICc

#Preparing Residual Plot

fit.rs=resid(fit)

plot(time(unemp), fit.rs,main="Residual plot") 
abline(0, 0) 

#Plotting acf and pacf
acf(resid(fit))
pacf(resid(fit))


```

##5.ARIMA model

Differencing the data and plotting it

```{r message=FALSE, echo=FALSE}
plot(unemp)
d_unemp <- diff(unemp)
plot(d_unemp)
```

Seasonally difference d_unemp and plot it

```{r message=FALSE, echo=FALSE}
d_unemp <- diff(unemp)
dd_unemp <- diff(d_unemp, lag = 12)  
plot(dd_unemp)
```
The data appaears to stationary


Ploting P/ACF pair of fully differenced data to lag 60


```{r message=FALSE, echo=FALSE,include=TRUE}
dd_unemp <- diff(diff(unemp), lag = 12)
acf2(dd_unemp)
sarima(unemp, p=2,d=1,q=0,P=0,D=1,Q=1,S=12)
```

##6.Model Diagnostics





