---
title: "Time Series Analysis - Project"
author: "Divya Chandrashekar"
date: "April 25, 2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error =TRUE)

```

Installing required packages
```{r}
install.packages("astsa")
install.packages("devtools")
devtools::install_github("nickpoison/astsa")

library(astsa)
```

##1. Data Description

a) Data Source:U.S. Unemployment. Astsa package. Dataset - unemp
 
```{r unemp}
unemp<-astsa::unemp
```

b) Variable Description

Monthly U.S. Unemployment series (1948-1978, n = 372)
```{r}
summary(unemp)
#checking class to confirm it is a ts
class(unemp)
```

c) ts description

```{r}
start(unemp)

end(unemp)

frequency(unemp)
```
From these results it can be see that the series starts with 1948 January and ends at 1978 December, once cycle is 12 months.


##2. Data Exploration

a) Plotting data

```{r message=FALSE, echo=FALSE}
plot(unemp, ylab = "Unemployment",main="US Unemployment over time")
```


```{r message=FALSE, echo=FALSE}
plot(unemp,ylab = "Unemployment",main="US Unemployment over time (with trend line")
abline(reg=lm(unemp~time(unemp))) # fit a trend line
```

For further EDA,
Box plot across months to explore seasonal effects 
(Also aggregating annualy)

```{r message=FALSE, echo=FALSE}
plot(aggregate(unemp), ylab = "Unemplyment Annual",main="Annal US Unemployment over time")
boxplot(unemp ~ cycle(unemp),ylab="US Unemployment",main="Boxplot of Unemployment across months")
```


From the preliminary plot of raw data, an increasing trend can be observed.
From the boxplot across months, highest unemployment is observed in June. 


b)Transforming data

Differencing the unemployment data and plotting it
```{r message=FALSE, echo=FALSE}
d_unemp <- diff(unemp)
plot(d_unemp, main ="Unemplyment data plot (difference)")
```

Seasonally differencing d_unemp and plotting it
```{r}
dd_unemp <- diff(d_unemp, lag = 12)  
plot(dd_unemp,main="Unemplyment data plot (seasonal difference)")
```


By differencing, the trend and seasonal variation in unemployment have been removed.
The series now appears to be stationary.

##3. Data Decomposition

Decomposing series into trend, seasonal and residual components
and plotting these

```{r message=FALSE, echo=FALSE}
plot(decompose(unemp))
u<-decompose(unemp)

trend<-u$trend
seasonal<-u$seasonal
residual<-u$random


plot(trend,ylab="US Unemployment",main="Plot of Unemployment trend")
plot(seasonal,ylab="US Unemployment",main="Plot of Unemployment seasonal")
plot(residual,ylab="US Unemployment",main="Plot of Unemployment residuals")

ts.plot(cbind(trend, trend+seasonal, main="Unemployment - trend, trend+seasonal"), lty=1:2)
```

Since the series shows an increasing trend, but seasonal effect does not, an additive decomposition model looks appropriate.
The residuals seem to show an increasing trend, so a transformation (log) can be considered.

##4.Regression

Regressing unemployment data on time
```{r}
fit <- lm(unemp~time(unemp), na.action=NULL)

summary(fit)

summary(aov(fit))
num = length(unemp)


AIC(fit)/num - log(2*pi) # AIC
BIC(fit)/num - log(2*pi) # BIC
(AICc = log(sum(resid(fit)^2)/num) + (num+5)/(num-5-2)) # AICc

#Preparing Residual Plot

fit.rs=resid(fit)

plot(time(unemp), fit.rs,main="Residual plot") 
abline(0, 0) 

#Plotting acf and pacf
acf(resid(fit))
pacf(resid(fit))


```

Interpretation of linear regression model:

The adjusted R square of the model is 0.49 the closer it gets to 1 better the model fits the data.Transformations such as log or taking a difference may result in a model with higher R squared value.

The slow linear decay of the ACF plot is typical of a non-stationary time series.

##5.ARIMA model

Differencing the data and plotting it

```{r message=FALSE, echo=FALSE}
plot(unemp, main="Unemployment")
d_unemp <- diff(unemp)
plot(d_unemp,"Unemployment (difference)")
```

Seasonally difference d_unemp and plot it

```{r message=FALSE, echo=FALSE}
d_unemp <- diff(unemp)
dd_unemp <- diff(d_unemp, lag = 12)  
plot(dd_unemp,main="Unemployment - seasonal difference")
```

The data appears to be stationary.Plotting P/ACF pair of fully differenced data to lag 60


```{r message=FALSE, echo=FALSE,include=TRUE}
dd_unemp <- diff(diff(unemp), lag = 12)
acf2(dd_unemp)
sarima(unemp, p=2,d=1,q=0,P=0,D=1,Q=1,S=12)
```

For the nonseasonal component:  PACF cuts off at lag 2 and the ACF tails off and in the seasonal component: the ACF cuts off at lag 12 and the PACF tails off at lags 12, 24, 36 etc.


##6.Model Diagnostics

Ploting P/ACF pair of fully differenced data to lag 60


```{r message=FALSE, echo=FALSE,include=TRUE}
dd_unemp <- diff(diff(unemp), lag = 12)
acf2(dd_unemp)
sarima(unemp, p=0,d=0,q=0,P=0,D=0,Q=1,S=12)

```
```{r message=FALSE, echo=FALSE,include=TRUE}
dd_unemp <- diff(diff(unemp), lag = 12)
acf2(dd_unemp)
sarima(unemp, p=1,d=1,q=0,P=0,D=0,Q=1,S=12)

```


```{r message=FALSE, echo=FALSE,include=TRUE}
dd_unemp <- diff(diff(unemp), lag = 12)
acf2(dd_unemp)
sarima(unemp, p=2,d=2,q=0,P=0,D=0,Q=1,S=12)

```

Based on the above models,the first arima model (p=2,d=1,q=0,P=0,D=1,Q=1,S=12) has the best fit.The AIC and BIC values for it are lowest (AIC=7.12457 and BIC=6.156174)
From this model, residuals are small in magnitude, appear to be uncorrelated close to normal and have a mean near zero.




